---
output:
  github_document:
    toc: TRUE
    toc_depth: 2
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = T,
  error = T
)
```



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = T)
options(tidyverse.quiet = TRUE)
```

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->



```{r, eval = F}
devtools::create(".")
```



> ... creating new stats is one of the most useful ways to extend the capabilities of ggplot2.' -- ggplot2: Elegant Graphics for Data Analysis  *Yes!*

Current methods for defining new user-facing stat and geom functions may be considered prohibitive for in-script, on-the-fly-use.   statexpress attempts to make concise definition possible - almost as concise as the pre-processing of data that you might have to do *in absence* of the proposed statexpress functionality.  With statexpress, extending ggplot2's capabilities by creating under-the-hood statisical transformation routines could happen *not only* as development activity, but also in data-analytic contexts.

Because creating new stats is so useful and can contribute so much to user's expressiveness w/ ggplot2 and because users are likely to intuit how Stats and Geoms work together at a high level, perhaps it's an activity that should hold a special place in the extension world - where not so much is expected of extenders as is suggested by some resources (e.g. Internals Chapter of ggplot2 which follows):

> *Maybe not?*  For a ggplot2 developer who hopes to design extensions, however, ... understanding [how ggplot2 translates this plot specification into an image] is paramount. (ggplot2 internals)

> ... the ease with which extensions can be made means that writing one-off extensions to solve a particular plotting challenge is also viable. *(More true with shortcut?)*


*Creating a geom_() [by defining a new Stat] from scratch is an involved process that doesn’t lend itself to simple transformations.* Elio Campitelli, in blog post 'How to make a generic stat in ggplot2' ^[Elio offers a solution, `stat_rasa()`, but it works to mimic just a small part of the functionality offered by creating new Stats.  {statexpress} aim is to allow full flexibility of designing a new Stat as with classical use, while keeping boilerplate at bay. It should be notated that a new `stat_manual()` function under development in base ggplot2 also achieves the limited goals Campitelli managed with `stat_rasa()`.]


A few approaches have been combine here.

1. `qstat()`...
2. `qlayer()`...



Let's go!

# `qstat`

A quick way to test out new Stats is to use them in combination with geom_*() user facing functions.  For example,

```{r}
library(tidyverse)
# Step 1. Define compute
compute_group_means <- function(data, scales){
 
  data %>%
    summarise(x = mean(x, na.rm = T),
              y = mean(y, na.rm = T))
 
}

# Step 2. Define Stat
StatMeans <- ggproto("StatMeans",
                     Stat,
                     compute_group = compute_group_means)

# Step 3. Use Stat
palmerpenguins::penguins %>%
  ggplot() +
  aes(bill_length_mm, bill_depth_mm) +
  geom_point() +
  geom_point(stat = StatMeans, size = 8)
```


But naming things is hard; and if we're just starting off, we might just want to have a way to check out our compute in a visual space kind of directly. So we introduce a way to test out our compute without a dedicated Stat in the global environment and without having to name it. That might look something like this two-step process:

```{r, eval = F}
# Step 1. Define compute
compute_group_means <- function(data, scales){
 
  data %>%
    summarise(x = mean(x, na.rm = T),
              y = mean(y, na.rm = T))
 
}

# Step 2. Use Compute
palmerpenguins::penguins %>%
  ggplot() +
  aes(bill_length_mm, bill_depth_mm) +
  geom_point() +
  geom_point(stat = qstat(compute_group_means), size = 8)
```

And you could start thinking about something user-facing like this:

```{r}
geom_means_point <- function(...){
  geom_point(stat = qstat(compute_group_means))
  }

geom_means <- geom_means_point

```

This might be enabled by a simple set of convenience functions qstat_group, qstat_panel, and qstat_layer, where are specified as follows.  qstat is just an alias of qstat_group (and can actually be used for panel or layer computation by spelling out compute_panel argument).

```{r}
knitrExtra:::chunk_to_r("a_qstat")
```

```{r a_qstat}
qstat <- function(compute_group = ggplot2::Stat$compute_group, ...) {

  ggproto("StatTemp", Stat, compute_group = compute_group, ...)
 
}

qstat_group <- function(compute_panel, ...) {

  ggproto("StatTemp", Stat, qstat_group = qstat_group, ...)
 
}

qstat_panel <- function(compute_panel, ...) {

  ggproto("StatTemp", Stat, compute_panel = compute_panel, ...)
 
}

qstat_layer <- function(compute_layer, ...) {

  ggproto("StatTemp", Stat, compute_layer = compute_layer, ...)
 
}
```

Basic example/test:

```{r}
palmerpenguins::penguins %>%
  ggplot() +
  aes(bill_length_mm, bill_depth_mm) +
  geom_point() +
  geom_point(stat = qstat(compute_group_means), size = 8)
```





# `qlayer()`

You might however, also need something like qlayer.  This can be useful if you want to 'try on' a user facing stat_*() function - in other words, don't lock in a geom.  Furthermore, you've got to use it in cases where user-facing geom functions don't give you the option of changing the Stat (no stat argument) as with geom_vline/GeomVline.

qlayer is essentially a flexible cross between geom_point and stat_identity.

```{r}
knitrExtra:::chunk_to_r("b_qlayer")
```

```{r b_qlayer}
qlayer <- function (mapping = NULL,
                    data = NULL,
                    stat = StatIdentity,
                    geom = GeomPoint,
                    position = position_identity(),
                    ...,
                    na.rm = FALSE,
                    show.legend = NA,
                    inherit.aes = TRUE)
{
  layer(
    data = data,
    mapping = mapping,
    stat = stat,
    geom = geom,
    position = position,
    show.legend = show.legend,
    inherit.aes = inherit.aes,
    params = rlang::list2(na.rm = na.rm, ...)
  )
}
```


```{r}
stat_means <- function(geom = GeomPoint, ...){
 
  qlayer(geom = geom, stat = qstat(compute_group_means), ...)
 
}

ggplot(cars) +
  aes(speed, dist) +
  geom_point() +
  stat_means(size = 12) +
  stat_means(geom = "label",
             label = "means x and y")

ggplot(palmerpenguins::penguins) +
  aes(bill_length_mm, bill_depth_mm) +
  geom_point() +
  aes(color = species) +
  stat_means(size = 12) +
  stat_means(geom = "label",
             aes(label = species))
```

Now looking at the case of GeomVline which doesn't have a user facing function that allows Stat to be input.

```{r}
compute_group_xmean <- function(data, scales){
 
  data %>% summarise(xintercept = mean(x))
 
}

geom_xmean_vline <- function(...){
 
  qlayer(stat = qstat(compute_group_xmean),
         geom = GeomVline, ...)
 
}


ggplot(faithful) +
  aes(waiting) +
  geom_histogram() +
  geom_xmean_vline()
```

Here's a good moment to show that qstat can also be used to more fully flesh out a hypothetical stat. Given the warning we might define a quick stat within the user-facing layer, so that we are warning-free.


```{r}
geom_xmean_vline <- function(...){
 
  QStat <- qstat(compute_group_xmean,
                 dropped_aes = "x")
 
  qlayer(stat = QStat,
         geom = GeomVline, ...)
 
}

ggplot(faithful) +
  aes(waiting) +
  geom_histogram() +
  geom_xmean_vline()
```



It is also possible of course to commit to a Stat, and use qlayer to speed your definitions along, if you're layer specification is anticipated to be a tad vanilla (as mine, for example, mostly are!).

```{r}
stat_means <- function(geom = "point", ...){
 
  qlayer(geom = geom, stat = StatMeans, ...)
 
}

ggplot(palmerpenguins::penguins) +
  aes(bill_length_mm, bill_depth_mm) +
  geom_point() +
  aes(color = species) +
  stat_means(size = 12) +
  stat_means(geom = "label",
             aes(label = species))




```

# `proto_update` and `qproto_update`

Some new ideas...

3. `qproto_update()` ... qproto_update(_inherit = GeomPoint, default_aes_update = aes(color = "blue")) 


4. `default_aes_update()` ...

```{r}
library(tidyverse)


proto_update <- function(`_class`, `_inherit`, default_aes_update = NULL, ...){
  
  if(!is.null(default_aes_update)){
  
    default_aes <- aes(!!!modifyList(`_inherit`$default_aes, default_aes_update))
    
    }
  
  ggproto(`_class`, `_inherit` = `_inherit`, default_aes = default_aes, ...)
  
}

GeomPointCornflowerblue <- proto_update("GeomPointCornflowerblue", 
                                    GeomPoint,
                                    default_aes_update = aes(color = "cornflowerblue"))

ggplot(cars) + 
  aes(speed, dist) + 
    layer(geom = GeomPointCornflowerblue,
          stat = StatIdentity,
          position = "identity")

qproto_update <- function(`_inherit`, default_aes_update = NULL, ...){
  
  proto_update("protoTemp", 
               `_inherit`, 
               default_aes_update = default_aes_update,
               ...)
  
}

ggplot(cars) + 
  aes(speed, dist) + 
    layer(geom = qproto_update(GeomPoint, aes(color = "blue")),
          stat = StatIdentity,
          position = "identity")


default_aes_update <- function(`_inherit`, update){
  
  aes(!!!modifyList(`_inherit`$default_aes, update))
  
}


GeomPointPink <- ggproto("GeomPointPink", GeomPoint, 
                         default_aes = 
                           default_aes_update(GeomPoint, aes(color = "pink")))

ggplot(cars) + 
  aes(speed, dist) + 
    layer(geom = GeomPointPink,
          stat = StatIdentity,
          position = "identity")






```



# a bunch of examples and ad-hoc tests...

## geom_means

test

```{r}
library(tidyverse)
group_means <- function(data, scales){
 
 data |>
    summarise(x = mean(x),
              y = mean(y))
 
}

geom_means <- function(...){
  geom_point(stat = qstat(group_means), ...)
}

mtcars |>
  ggplot() +
  aes(x = wt,
      y = mpg) +
  geom_point() +
  geom_means(size = 6)

last_plot() +
  aes(color = factor(cyl))

mtcars |>
  ggplot() +
  aes(x = wt,
      y = mpg) +
  aes(color = factor(cyl)) +
  geom_point(show.legend = F) +
  geom_means(size = 6, show.legend = F)

mtcars |>
  ggplot() +
  aes(x = wt,
      y = mpg) +
  geom_point() +
  geom_means(size = 6, aes(color = NULL)) +
  aes(color = factor(cyl))
```


## geom_center_label

```{r}
group_label_at_center <- function(data, scales){
 
 data %>%
    group_by(label) %>%
    summarise(x = mean(x, na.rm = T),
              y = mean(y, na.rm = T))
 
}

geom_center_label <- function(...){
  geom_label(stat = qstat(group_label_at_center), ...)
}


palmerpenguins::penguins |>
  ggplot() +
  aes(x = bill_length_mm,
      y = bill_depth_mm) +
  geom_point() +
  aes(label = "All") +
  geom_center_label()


last_plot() +
  aes(color = species, label = species)


geom_center_text <- function(...){
  geom_text(stat = qstat(group_label_at_center), ...)
}


palmerpenguins::penguins |>
  ggplot() +
  aes(x = bill_length_mm,
      y = bill_depth_mm) +
  geom_point(aes(color = bill_length_mm)) +
  aes(color = species) +
  geom_center_text(color = "Black",
    aes(label = species),
                     
                    alpha = .8,
                   size = 5,
                   fontface = "bold")

layer_data(i = 2)
```

## geom_post

```{r}
compute_post <- function(data, scales){
 
  data %>%
    mutate(xend = x,
           yend = 0)
 
}

geom_post <- function(...){
  geom_segment(stat = qstat(compute_post), ...)
}


data.frame(outcome = 0:1, prob = c(.4, .6)) |>
  ggplot() +
  aes(x = outcome,
      y = prob) +
  geom_post() +
  geom_point() +
  labs(title = "probability by outcome")


```

## geom_xmean

```{r}
compute_xmean <- function(data, scales){
 
  data %>%
    summarize(xintercept = mean(x))
 
}

geom_xmean <- function(...){
  qlayer(stat = qstat(compute_xmean,
                      dropped_aes = c("x", "y")),
         geom = "vline", ...)
}

mtcars |>
  ggplot() +
  aes(x = wt,
      y = mpg) +
  geom_point() +
  geom_xmean(linetype = "dashed")

last_plot() +
  aes(color = factor(cyl))
```

## geom_quantile

test of passing arguments from compute

```{r}
compute_xy_quantile <- function(data, scales, q = .5){
 
  data %>%
    summarise(x = quantile(x, q),
              y = quantile(y, q))
 
}

geom_quantile <- function(...){
  geom_point(stat = qstat(compute_xy_quantile), ...)  #adding point means that aes is automatically figured out! ok
}

mtcars |>
  ggplot() +
  aes(x = wt,
      y = mpg) +
  geom_point() +
  geom_quantile(size = 8, aes(color = "q = .5"), q = .5) +
  geom_quantile(size = 8, q = 1, aes(color = "q =  1")) +
  geom_quantile(aes(color = "q =  .9"), size = 8, q = .9)

```


## `geom_highlight()`

```{r, eval = T}
compute_panel_highlight <- function(data, scales, which_id = NULL){

data %>%
  arrange(highlight_condition) %>%
  mutate(group = fct_inorder(grouping))
 
}

stat_highlight <- function(geom = "line", ...){
 
  qlayer(stat = qstat_panel(compute_panel_highlight,
                      default_aes =
                        aes(color = after_stat(highlight_condition))),
             geom = geom,
             ...)
}

gapminder::gapminder %>%
  filter(continent == "Americas") %>%
  ggplot() +
  aes(x = year, y = lifeExp,
      grouping = country,
      highlight_condition =
        country == "Bolivia") +
  stat_highlight(linewidth = 3)

gapminder::gapminder %>%
  filter(year == 2002) %>%
  ggplot() +
  aes(x = gdpPercap, y = lifeExp,
      grouping = continent,
      highlight_condition =
        continent == "Europe") +
  stat_highlight(geom = "point", show.legend = F) +
  scale_color_manual(values = c("grey", "darkolivegreen"))

```


# One-liners?

Since were organize with variable function input in first position and geom in section position, and we can do one-liners (or two) use positioning for arguments.




## `geom_xmean_line()` in 137 characters

```{r}
library(tidyverse)

geom_xmean_line <- function(...){qlayer(stat = qstat(function(data, scales) data |> summarize(xintercept = mean(x)) , dropped_aes = c("x", "y")), geom =  "vline", ...)}


ggplot(cars) +
  aes(speed, dist) +
  geom_point() +
  geom_xmean_line(linetype = 'dashed')
 
last_plot() +
  aes(color = dist > 50)

ggplot(cars) +
  aes(speed, dist) +
  geom_point() +
  geom_xmean_line(linetype = 'dashed',
                  data = . %>% tail,
                  aes(color = dist > 50))

ggplot(cars) +
  aes(speed, dist) +
  geom_point() +
  geom_xmean_line(data = . %>% filter(speed < 10))
```

## `geom_xmean` in 99 characters

```{r}
geom_xmean <- function(...){geom_point(stat = qstat(function(data, scales){data |> summarize(x = mean(x), y = I(.025))}), ...)}

ggplot(cars) +
  aes(speed, dist) +
  geom_point() +
  geom_xmean(size = 8, shape = "diamond")

last_plot() +
  aes(color = dist > 50)
```


## `geom_post()` in 101 characters, `stat_expectedvalue()` in 113, `geom_expectedvalue_label()` 171...

May I buy a visually enhanced probability lesson for 400 characters? Yes please.

```{r cars}
geom_post <- function(...){geom_segment(stat = qstat(function(data, scales){data |> mutate(xend = x, yend = 0)}), ...)}

data.frame(prob = c(.4,.6), outcome = c(0, 1)) %>%
ggplot(data = .) +
  aes(outcome, prob) +
  geom_post() +
  geom_point()

stat_expectedvalue <- function(geom = "point", ...){
 
  qlayer(stat = qstat(function(data, scales){data |> summarise(x = sum(x*y), y = 0)},
                      default_aes = aes(label = after_stat(round(x, 2)))),
             geom = geom,
             ...)
 
  } # point is default geom

last_plot() +
  stat_expectedvalue()

last_plot() +
  stat_expectedvalue(geom = "text", vjust = 0) +
  stat_expectedvalue(geom = "text", 
                     label = "The Expected Value",
                     vjust = 1)
```



### `geom_proportion()` and  `geom_proportion_label()`


```{r}
rep(1, 15) |>
  c(0) %>%
  data.frame(outcome = .) |>
  ggplot() +
  aes(x = outcome) +
  geom_dotplot()


geom_proportion <- function(...){geom_point(stat = qstat_panel(function(data, scales) data |> summarise(x = sum(x)/length(x), y = 0)), ...)}   # this should work for T/F too when rasa_p is in play


last_plot() +
  geom_proportion(shape = "triangle", size = 5)

geom_proportion_label <- function(...){geom_text(stat = qstat_panel(function(data, scales) data |> summarise(x = sum(x)/length(x), y = 0) |> mutate(label = round(x,2))), vjust = 0, ...)}   # this should work for T/F too when rasa_p is in play

last_plot() +
  geom_proportion_label()

# last_plot() +
#   geom_proportion_label() +
#   ggsample::facet_bootstrap()
#
# layer_data(i = 2)
#
#
#
# rep(0:1, 10000) %>% # very large 50/50 sample
#   data.frame(outcome = .) |>
#   ggplot() +
#   aes(x = outcome) +
#   geom_dotplot() +
#   geom_proportion() +
#   geom_proportion_label() +
#   ggsample::facet_sample(n_facets = 25,
#     n_sampled = 16) ->
# p; p
#  
#
# layer_data(p, i = 2) |>
#   ggplot() +
#   aes(x = x) +
#   geom_rug() +
#   geom_histogram()


```


## `geom_means` in 131 characters

```{r}
geom_means <- function(...){geom_point(stat = qstat(function(data, scales) data |> summarize(x = mean(x, na.rm = T), y = mean(y, na.rm = T))), ...)}

palmerpenguins::penguins %>%
  ggplot() +
  aes(bill_length_mm, bill_depth_mm) +
  geom_point() +
  geom_means(size = 7)

last_plot() +
  geom_means(size = 8, shape = "diamond", aes(color = bill_depth_mm > 17))  # why does aes work here w/o mapping =


last_plot() +
  geom_means(size = 8,
             shape = "square",
             data = . %>% filter(species == "Chinstrap"),
             aes(color = bill_depth_mm > 17),
             alpha = .6)
```

## `geom_grouplabel_at_means()`

```{r}
geom_grouplabel_at_means <-  function(...){geom_label(stat = qstat(function(data, scales) data |> group_by(label) |> summarize(x = mean(x, na.rm = T), y = mean(y, na.rm = T))), ...)}

palmerpenguins::penguins %>%
  ggplot() +
  aes(bill_length_mm, bill_depth_mm, label = species, group = species) +
  geom_point() +
  geom_grouplabel_at_means(size = 7)
```





# Part II. Packaging and documentation  🚧 ✅

## Phase 1. Minimal working package

To build a minimal working package, 1) we'll need to document dependencies 2) send functions to .R files in the R folder, 3) do a package check (this will A. Document our functions and B. this will help us identify problems - for example if we've failed to declare a dependency) and 4) install the package locally.  

Then we'll write up a quick advertisement for what our package is able to do in the 'traditional readme' section.  This gives us a chance to test out the package.

```{r, eval = F}
### Bit 2a: in the function(s) you wrote above make sure dependencies to functions using '::' syntax to pkg functions
usethis::use_package("ggplot2") # Bit 2b: document dependencies, w hypothetical ggplot2
```



```{r, eval = F}
# Bit 4: document functions and check that package is minimally viable
devtools::check(pkg = ".")  
```


```{r, eval = F}
# Bit 5: install package locally
devtools::install(pkg = ".", upgrade = "never")
```


### Bit 7. Write traditional README that uses built package (also serves as a test of build). 🚧 ✅

The goal of the {statexpress} package is to ...

Install package with:

```
remotes::install_github("EvaMaeRey/statexpress")
```

Once functions are exported you can remove go to two colons, and when things are are really finalized, then go without colons (and rearrange your readme...)

```{r, eval = T}
rm(list = ls())
library(statexpress)
library(ggplot2)
library(tidyverse)

compute_group_balancing = function(data, scales){
    data |>
      summarize(x = mean(x), y = I(.05))
    }

geom_xbalance <- function(...){
 
  geom_point(stat = statexpress:::qstat(compute_group_balancing), ...)
 
  }

ggplot(cars) +
  aes(speed, dist) +
  geom_point() +
  geom_xbalance(size = 8, shape = "diamond")

last_plot() +
  aes(color = dist > 50)

last_plot() +
  geom_xbalance(mapping = aes(size = dist > 75))

```

```{r files-report-and-stop-knit, eval = T}
# for quick knit (exiting early) change eval to TRUE
fs::dir_tree(recurse = T)
knitr::knit_exit()
```

# Bit 8: Compile readme

# Bit 9: Push to github

# Bit 10: listen and iterate

## Phase 3: Settling and testing 🚧 ✅

### Bit A. Added a description and author information in the [DESCRIPTION file](https://r-pkgs.org/description.html) 🚧 ✅

### Bit B. Added [roxygen skeleton](https://r-pkgs.org/man.html) for exported functions. 🚧 ✅

### Bit D. Settle on [examples](https://r-pkgs.org/man.html#sec-man-examples).  Put them in the roxygen skeleton and readme. 🚧 ✅

### Bit C. Chosen a [license](https://r-pkgs.org/license.html)? 🚧 ✅

```{r, eval = F}
usethis::use_mit_license()
```

### Bit D. Use life-cycle badge

```{r, eval = F}
usethis::use_lifecycle_badge("experimental")
```


### Bit E. Written formal [tests](https://r-pkgs.org/testing-basics.html) of functions and save to test that folders 🚧 ✅




That would look like this...

```{r test_calc_times_two_works, eval = F}
library(testthat)

test_that("calc times 2 works", {
  expect_equal(times_two(4), 8)
  expect_equal(times_two(5), 10)
 
})
```


```{r, eval = F}
knitrExtra::chunk_to_tests_testthat("test_calc_times_two_works")
```




### Bit F. Check again. Addressed notes, warnings and errors. 🚧 ✅


```{r, eval = F}
devtools::check(pkg = ".")
```

## Phase 4. Promote to wider audience...  🚧 ✅

### Bit A. Package website built? 🚧 ✅


### Bit B. Package website deployed? 🚧 ✅

## Phase 5: Harden/commit: Submit to CRAN/RUniverse 🚧 ✅

# Appendix: Reports, Environment

## Description file complete?  🚧 ✅

```{r, eval = F}
readLines("DESCRIPTION")
```

## Environment  🚧 ✅

Here I just want to print the packages and the versions

```{r}
all <- sessionInfo() |> print() |> capture.output()
all[11:17]
```

## `devtools::check()` report

```{r, eval=F, error = T, results="hide", warning=F}
devtools::check(pkg = ".")
```

## Package directory file tree

```{r}
fs::dir_tree(recurse = T)
```
