---
output: 
  github_document:
    toc: TRUE
    toc_depth: 2
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = T,
  error = T
)
```



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = T)
options(tidyverse.quiet = TRUE)
```




```{r, eval = F}
devtools::create(".")
```



> ... creating new stats is one of the most useful ways to extend the capabilities of ggplot2.' -- ggplot2: Elegant Graphics for Data Analysis  *Yes!*

Current methods for defining new user-facing stat and geom functions may be considered prohibitive for in-script, on-the-fly-use.   statexpress attempts to make concise definition possible - almost as concise as the pre-processing of data that you might have to do *in absence* of the proposed statexpress functionality.  With statexpress, extending ggplot2's capabilities by creating under-the-hood statisical transformation routines could happen *not only* as development activity, but also in data-analytic contexts.

Because creating new stats is so useful and can contribute so much to user's expressiveness w/ ggplot2 and because users are likely to intuit how Stats and Geoms work together at a high level, perhaps it's an activity that should hold a special place in the extension world - where not so much is expected of extenders as is suggested by some resources (e.g. Internals Chapter of ggplot2 which follows):

> *Maybe not?*  For a ggplot2 developer who hopes to design extensions, however, ... understanding [how ggplot2 translates this plot specification into an image] is paramount. (ggplot2 internals)

> ... the ease with which extensions can be made means that writing one-off extensions to solve a particular plotting challenge is also viable. *(More true with shortcut?)*


A few approaches have been combine here. 

1. `qstat()`...
2. `qlayer()`...

Let's go!

# `qstat`



```{r}
knitrExtra:::chunk_to_r("a_qstat")
```

```{r a_qstat}
qstat <- function(compute_group = ggplot2::Stat$compute_group, ...) {

  ggproto("StatTemp", Stat, compute_group = compute_group, ...)
  
}
```

# then define user-facing `qlayer()`, which gives the presets of geom_point and stat_identity.

```{r}
knitrExtra:::chunk_to_r("b_qlayer")
```

```{r b_qlayer}
qlayer <- function (mapping = NULL, data = NULL, stat = "identity", geom = "point", position = "identity", 
    ..., na.rm = FALSE, show.legend = NA, inherit.aes = TRUE) 
{
    layer(data = data, mapping = mapping, stat = stat, geom = geom, 
        position = position, show.legend = show.legend, inherit.aes = inherit.aes, 
        params = rlang::list2(na.rm = na.rm, ...))
}
```


# maybe something sf-facing


## geom_means

```{r}
library(tidyverse)
group_means <- function(data, scales){
  
 data |> 
    summarise(x = mean(x),
              y = mean(y))
  
}

geom_means <- function(...){
  geom_point(stat = qstat(group_means), ...)
}

mtcars |>
  ggplot() + 
  aes(x = wt,
      y = mpg) + 
  geom_point() + 
  geom_means(size = 6)

last_plot() + 
  aes(color = factor(cyl))
```


## geom_center_label

```{r}
group_label_at_center <- function(data, scales){
  
 data %>% 
    group_by(label) %>% 
    summarise(x = mean(x, na.rm = T),
              y = mean(y, na.rm = T))
  
}

geom_center_label <- function(...){
  geom_label(stat = qstat(group_label_at_center), ...)
}


palmerpenguins::penguins |>
  ggplot() +
  aes(x = bill_length_mm,
      y = bill_depth_mm) +
  geom_point() +
  aes(label = "All") +
  geom_center_label()


last_plot() +
  aes(color = species, label = species)


geom_center_text <- function(...){
  geom_text(stat = qstat(group_label_at_center), ...)
}


palmerpenguins::penguins |>
  ggplot() +
  aes(x = bill_length_mm,
      y = bill_depth_mm) +
  geom_point(aes(color = bill_length_mm)) +
  aes(color = species) +
  geom_center_text(color = "Black",
    aes(label = species), 
                     
                    alpha = .8,
                   size = 5, 
                   fontface = "bold")

layer_data(i = 2)
```

## geom_post

```{r}
compute_post <- function(data, scales){
  
  data %>% 
    mutate(xend = x,
           yend = 0)
  
}

geom_post <- function(...){
  geom_segment(stat = qstat(compute_post), ...)
}


data.frame(outcome = 0:1, prob = c(.4, .6)) |>
  ggplot() + 
  aes(x = outcome,
      y = prob) + 
  geom_post() + 
  geom_point() + 
  labs(title = "probability by outcome")


```

## geom_xmean

```{r}
compute_xmean <- function(data, scales){
  
  data %>% 
    summarize(xintercept = mean(x))
  
}

geom_xmean <- function(...){
  qlayer(stat = qstat(compute_xmean, 
                      dropped_aes = c("x", "y")), 
         geom = "vline", ...)
}

mtcars |>
  ggplot() + 
  aes(x = wt,
      y = mpg) + 
  geom_point() + 
  geom_xmean(linetype = "dashed")

last_plot() + 
  aes(color = factor(cyl))
```

## geom_quantile

```{r}
compute_xy_quantile <- function(data, scales, q = .5){
  
  data %>% 
    summarise(x = quantile(x, q),
              y = quantile(y, q)) 
  
}

geom_quantile <- function(...){
  geom_point(stat = qstat(compute_xy_quantile), ...)  #adding point means that aes is automatically figured out! ok
}

mtcars |>
  ggplot() +
  aes(x = wt,
      y = mpg) +
  geom_point() +
  geom_quantile(size = 8, aes(color = "q = .5"), q = .5) +
  geom_quantile(size = 8, q = 1, aes(color = "q =  1")) +
  geom_quantile(aes(color = "q =  .9"), size = 8, q = .9)

```


## `geom_highlight()`

```{r, eval = T}
compute_panel_highlight <- function(data, scales, which_id = NULL){

data %>% 
  arrange(highlight_condition) %>%
  mutate(group = fct_inorder(grouping))
  
}

stat_highlight <- function(geom = "line", ...){
  
  qlayer(stat = qstat(compute_panel = compute_panel_highlight, 
                      default_aes = 
                        aes(color = after_stat(highlight_condition))), 
             geom = geom, 
             ...)
}

gapminder::gapminder %>% 
  filter(continent == "Americas") %>% 
  ggplot() + 
  aes(x = year, y = lifeExp, 
      grouping = country, 
      highlight_condition = 
        country == "Bolivia") + 
  stat_highlight(linewidth = 3)

gapminder::gapminder %>% 
  filter(year == 2002) %>% 
  ggplot() + 
  aes(x = gdpPercap, y = lifeExp, 
      grouping = continent, 
      highlight_condition = 
        continent == "Europe") + 
  stat_highlight(geom = "point") + 
  scale_x_log10() + 
  scale_color_manual(values = c("grey", "darkolivegreen"))

```


# One-liners?

Since were organize with variable function input in first position and geom in section position, and we can do one-liners (or two) use positioning for arguments.




## `geom_xmean_line()` in 137 characters

```{r}
library(tidyverse)

geom_xmean_line <- function(...){qlayer(stat = qstat(compute_group = function(data, scales){data |> summarize(xintercept = mean(x))}, dropped_aes = c("x", "y")), geom =  "vline", ...)}


ggplot(cars) +
  aes(speed, dist) + 
  geom_point() + 
  geom_xmean_line(linetype = 'dashed')
  
last_plot() + 
  aes(color = dist > 50)

ggplot(cars) +
  aes(speed, dist) + 
  geom_point() + 
  geom_xmean_line(linetype = 'dashed', 
                  data = . %>% tail,
                  aes(color = dist > 50))

ggplot(cars) +
  aes(speed, dist) + 
  geom_point() + 
  geom_xmean_line(data = . %>% filter(speed < 10))
```

## `geom_xmean` in 99 characters

```{r}
geom_xmean <- function(...){geom_point(stat = qstat(function(data, scales){data |> summarize(x = mean(x), y = I(.025))}), ...)}

ggplot(cars) +
  aes(speed, dist) + 
  geom_point() + 
  geom_xmean(size = 8, shape = "diamond") 

last_plot() + 
  aes(color = dist > 50)
```


## `geom_post()` in 101 characters, `stat_expectedvalue()` in 113, `geom_expectedvalue_label()` 171... 

May I buy a visually enhanced probability lesson for 400 characters? Yes please.

```{r cars}
geom_post <- function(...){geom_segment(stat = qstat(function(data,scales){data |> mutate(xend = x, yend = 0)}), ...)}

data.frame(prob = c(.4,.6), outcome = c(0, 1)) %>% 
ggplot(data = .) +
  aes(outcome, prob) + 
  geom_post() +
  geom_point() 

stat_expectedvalue <- function(geom = "point", ...){
  
  qlayer(stat = qstat(function(data, scales){data |> summarise(x = sum(x*y), y = 0)},
                      default_aes = aes(label = after_stat(round(x, 2)))), 
             geom = geom, 
             ...)
  
  } # point is default geom

last_plot() + 
  stat_expectedvalue()

last_plot() +
  stat_expectedvalue(geom = "text", vjust = 0) + 
  stat_expectedvalue(geom = "text", label = "The Expected Value",
                     vjust = 1)
```

```{r}
knitr::knit_exit()

```

### `geom_proportion()` and  `geom_proportion_label()`


```{r}
rep(1, 15) |> 
  c(0) %>% 
  data.frame(outcome = .) |>
  ggplot() + 
  aes(x = outcome) + 
  geom_dotplot()

geom_proportion <- function(...){stat_panel(function(df) df |> summarise(x = sum(x)/length(x), y = 0), ...)}   # this should work for T/F too when rasa_p is in play

last_plot() + 
  geom_proportion(shape = "triangle")

geom_proportion_label <- function(...){stat_panel(function(df) df |> summarise(x = sum(x)/length(x), y = 0) |> mutate(label = round(x,2)), vjust = 0, "text", ...)}   # this should work for T/F too when rasa_p is in play

last_plot() + 
  geom_proportion_label()

# last_plot() + 
#   geom_proportion_label() + 
#   ggsample::facet_bootstrap()
# 
# layer_data(i = 2)
# 
# 
# 
# rep(0:1, 10000) %>% # very large 50/50 sample
#   data.frame(outcome = .) |>
#   ggplot() + 
#   aes(x = outcome) + 
#   geom_dotplot() +
#   geom_proportion() + 
#   geom_proportion_label() + 
#   ggsample::facet_sample(n_facets = 25,
#     n_sampled = 16) ->
# p; p 
#   
# 
# layer_data(p, i = 2) |>
#   ggplot() + 
#   aes(x = x) + 
#   geom_rug() + 
#   geom_histogram()


```


## `geom_means` in 131 characters

```{r}
geom_means <- function(...){stat_group(function(df) df |> summarize(x = mean(x, na.rm = T), y = mean(y, na.rm = T)), "point", ...)}

palmerpenguins::penguins %>% 
  ggplot() + 
  aes(bill_length_mm, bill_depth_mm) + 
  geom_point() + 
  geom_means(size = 7)

last_plot() +
  geom_means(size = 8, shape = "diamond", aes(color = bill_depth_mm > 17))  # why does aes work here w/o mapping =


last_plot() +
  geom_means(size = 8, 
             shape = "square", 
             data = . %>% filter(species == "Chinstrap"),
             aes(color = bill_depth_mm > 17),
             alpha = .6) 
```

## `geom_grouplabel_at_means()`

```{r}
geom_grouplabel_at_means <-  function(...){stat_group(function(df) df |> group_by(label) |> summarize(x = mean(x, na.rm = T), y = mean(y, na.rm = T)), "label", ...)}

palmerpenguins::penguins %>% 
  ggplot() + 
  aes(bill_length_mm, bill_depth_mm, label = species, group = species) + 
  geom_point() + 
  geom_grouplabel_at_means(size = 7)
```



# Then using stat_panel_sf w/ helper package sf2stat

```{r, warning = F, message = F, results='hide'}
nc <- sf::st_read(system.file("shape/nc.shp", package="sf"))
```

```{r}
geo_reference_nc <- nc |>
  dplyr::select(county_name = NAME, fips = FIPS) |>
  sf2stat:::sf_df_prep_for_stat(id_col_name = "county_name")


stat_nc_counties <- function(geom = "sf", ...){
  
  stat_panel_sf(geo_ref_data = geo_reference_nc, 
                fun = NULL, 
                geom = geom, 
                default_aes = 
                  ggplot2::aes(label = 
                                 after_stat(id_col)),
                ...)
  
}

nc %>% 
  sf::st_drop_geometry() %>% 
ggplot() + 
  aes(fips = FIPS) +
  stat_nc_counties() + 
  stat_nc_counties(geom = "text", check_overlap = T)

nc %>% 
  sf::st_drop_geometry() %>% 
ggplot() + 
  aes(fips = FIPS, fill = SID74) +
  stat_nc_counties() + 
  stat_nc_counties(geom = "text", 
                   data = . %>% head(),
                   aes(label = SID74),
                   color = "grey85") 
```





# Part II. Packaging and documentation  🚧 ✅ 

## Phase 1. Minimal working package

To build a minimal working package, 1) we'll need to document dependencies 2) send functions to .R files in the R folder, 3) do a package check (this will A. Document our functions and B. this will help us identify problems - for example if we've failed to declare a dependency) and 4) install the package locally.  

Then we'll write up a quick advertisement for what our package is able to do in the 'traditional readme' section.  This gives us a chance to test out the package.

```{r, eval = F}
### Bit 2a: in the function(s) you wrote above make sure dependencies to functions using '::' syntax to pkg functions 
usethis::use_package("ggplot2") # Bit 2b: document dependencies, w hypothetical ggplot2
```



```{r, eval = F}
# Bit 4: document functions and check that package is minimally viable
devtools::check(pkg = ".")  
```


```{r}
# Bit 5: install package locally
devtools::install(pkg = ".", upgrade = "never") 
```


### Bit 7. Write traditional README that uses built package (also serves as a test of build). 🚧 ✅ 

The goal of the {statExpress} package is to ...

Install package with:

```
remotes::install_github("EvaMaeRey/statExpress")
```

Once functions are exported you can remove go to two colons, and when things are are really finalized, then go without colons (and rearrange your readme...)

```{r, eval = T}
rm(list = ls())
library(statexpress)
library(ggplot2)
library(tidyverse)

geom_xmean <- function(...){stat_group(function(df) df |> summarize(x = mean(x), y = I(.05)), ...)}

ggplot(cars) +
  aes(speed, dist) + 
  geom_point() + 
  geom_xmean(size = 8, shape = "diamond") 

last_plot() + 
  aes(color = dist > 50)

last_plot() + 
  geom_xmean(mapping = aes(size = dist > 75))

```

```{r files-report-and-stop-knit, eval = T}
# for quick knit (exiting early) change eval to TRUE
fs::dir_tree(recurse = T)
knitr::knit_exit()
```

# Bit 8: Compile readme

# Bit 9: Push to github

# Bit 10: listen and iterate

## Phase 3: Settling and testing 🚧 ✅ 

### Bit A. Added a description and author information in the [DESCRIPTION file](https://r-pkgs.org/description.html) 🚧 ✅

### Bit B. Added [roxygen skeleton](https://r-pkgs.org/man.html) for exported functions. 🚧 ✅

### Bit D. Settle on [examples](https://r-pkgs.org/man.html#sec-man-examples).  Put them in the roxygen skeleton and readme. 🚧 ✅

### Bit C. Chosen a [license](https://r-pkgs.org/license.html)? 🚧 ✅

```{r, eval = F}
usethis::use_mit_license()
```

### Bit D. Use life-cycle badge

```{r, eval = F}
usethis::use_lifecycle_badge("experimental") 
```


### Bit E. Written formal [tests](https://r-pkgs.org/testing-basics.html) of functions and save to test that folders 🚧 ✅




That would look like this...

```{r test_calc_times_two_works, eval = F}
library(testthat)

test_that("calc times 2 works", {
  expect_equal(times_two(4), 8)
  expect_equal(times_two(5), 10)
  
})
```


```{r, eval = F}
knitrExtra::chunk_to_tests_testthat("test_calc_times_two_works")
```




### Bit F. Check again. Addressed notes, warnings and errors. 🚧 ✅


```{r, eval = F}
devtools::check(pkg = ".")
```

## Phase 4. Promote to wider audience...  🚧 ✅ 

### Bit A. Package website built? 🚧 ✅


### Bit B. Package website deployed? 🚧 ✅

## Phase 5: Harden/commit: Submit to CRAN/RUniverse 🚧 ✅ 

# Appendix: Reports, Environment

## Description file complete?  🚧 ✅

```{r, eval = F}
readLines("DESCRIPTION")
```

## Environment  🚧 ✅

Here I just want to print the packages and the versions

```{r}
all <- sessionInfo() |> print() |> capture.output()
all[11:17]
```

## `devtools::check()` report

```{r, eval=F, error = T, results="hide", warning=F}
devtools::check(pkg = ".")
```

## Package directory file tree

```{r}
fs::dir_tree(recurse = T)
```



